Modular Model-Checking of a Byzantine Fault-Tolerant Protocol

### 2016-12-08 doc/om-paper/paper.tex
final-ish proof read of sections 1-3

\newcommand{\OM}[1]{\ensuremath{\mathrm{OM}(#1)}\xspace}
\newcommand{\OMH}{\ensuremath{\mathrm{OMH}}\xspace}
\newcommand{\Id}{\ensuremath{\mathrm{Id}}\xspace}
\newcommand{\Msg}{\ensuremath{\mathrm{Msg}}\xspace}
\newcommand{\ERR}{\ensuremath{\mathrm{ERR}}\xspace}
\newcommand{\TO}{\ensuremath{\bm{\mathrm{T}}}\xspace}
\newcommand{\tld}[1]{#1_{*}\xspace}

@@

\begin{abstract}
With proof techniques like IC3 and $k$-induction, model-checking scales further
than ever before.  Still, fault-tolerant distributed systems are particularly
challenging to model-check given their large state spaces and
non-determinism. The typical approach to controlling complexity is to construct
ad-hoc abstractions of faults, message-passing, and behaviors.  However, these
abstractions come at the price of divorcing the model from its implementation
and making refactoring difficult. In this work, we present a framework for
fault-tolerant distributed system verification that combining ideas from the literature including calendar automata,
symbolic fault injection, and abstract transition systems, and then we apply the framework to model-checking various implementations of the Hybrid Oral Messages algorithm that differ in the fault model, timing model, and local node behavior. We show that despite being implementation-level models, the verifications are scalable and modular, insofar as isolated changes to an implementation require modular changes to the model and proofs.
This work is carried out in SAL model-checker.
\end{abstract}

@@

In this paper, we present a framework for modeling fault-tolerant distributed systems, and use that framework to verify several variant implementations of the Byzantine fault-tolerant Hybrid Oral Messages algorithm (\OMH)~\cite{Lincoln-Rushby}. The framework combines ideas from the literature and folklore to build scalable and modular formal models suitable for infinite-state model-checking, and it reduces the need for ad-hoc abstractions and optimizations while maintaining scalability. In Section~\ref{sec:model}, we present the important aspects of the framework, including calendar automata, originally developed by Dutertre and Sorea~\cite{cal}, symbolic fault-injection, and abstract transition systems for verification.

@@

We use the framework to verify implementation-level models of \OMH in which message passing is explicit, nodes are not enforced to execute strictly synchronously, and voting is explicit. In short, the models corresponds closely with an implementation of the algorithm. In Section~\ref{sec:byz}, we first describe \OMH, then an implementation of it that uses the Boyer-Moore Fast Majority Vote algorithm (Fast~MJRT)~\cite{mjrty}. We then describe a set of modular invariants, such that the invariants only concern specific aspects of the model (e.g., faults, local node behavior, or the passage of time). The verification is interesting in its own right, as it is the first fully parametric (on the number of nodes) model-checked \emph{implementation} of the algorithm.

@@

% ------------------------------------------------------------
\section{Formal Framework}\label{sec:model}
Here we describe our formal framework specialized for fault-tolerant distributed systems. The framework draws on three principal abstractions: calendar automata, symbolic fault injection, and abstract transition sytems; we describe each below.

@@

Real-time system verification in general-purpose model-checkers requires an explicit formalism of real-time progression. Trying to encode real-time clocks directly is difficult; in particular, one must avoid Zeno's paradox in which no progress is made because state transitions simply update real-valued variables by an infinitesimally small amount. To avoid this problem, Dutetre and Sorea developed \emph{calendar automata}~\cite{cal}, which is itself inspired by event calendars used in discrete-event simulation. Rather than encoding ``how much time has passed since the last event'', it encodes ``how far into the future is the next scheduled event'', and a real-valued variable representing the current time is updated to the next event time.

@@

Define a set of \emph{events} $e_0, e_1, \ldots, e_n \in E$. For now, we do not define events; intuitively, an event is a set of state variables (shortly, we will associate events with messages sent in a distributed system). When an event is \emph{enabled}, the transitions over events are enabled; otherwise, the variables stutter and maintain the same value.

@@

Let $cal$ be an event calendar and $c_i, c_j \in c$ be calendar events. Define an ordering on event calendars such that $c_i \leq c_j$ iff $t_i \leq t_j$, and $min(c) = \{ c_i | \forall c_j \in cal, \, c_i \leq c_j  \}$ are the minimum elements of $cal$.

Let a transition system $\mathcal{M} = (S, I, \rightarrow)$, be a set of states $S$, a set of initial states $I \subseteq S$, and a transition relation $\rightarrow \subseteq S \times S$. We implicitly assume a set of state variables such that a state $\sigma \in S$ is a total function that maps state variables to values. We distinguish two special state variables in a transition system: (1) $now \in \mathbb{R}$ denotes the current time in the state, and (2) $cal$ is an event calendar.

@@

\item \label{cal:c} In all states, if the current time is strictly less than every calendar event, then the only enabled transition is a \emph{time progress} update: $\forall \sigma \in S$, $\forall (e_i, t_i) \in \sigma(cal)$, if $\sigma(now) < t_i$, then $\forall \sigma'$ such that $\sigma \rightarrow \sigma'$, $\sigma' = \sigma[now := min(c)]$.

@@

Assume processes are indexed from a finite set $\Id \subset \mathbb{N}$. A \emph{channel} from process $i$ to $j$ is an ordered pair $(i,j)$. Define a set of messages $\Msg$. Given a channel and a timeout, let $send$ be a relation on messages sent on a channel at a given time:
$$send \subseteq \Id \times \Id \times \mathbb{R} \times \Msg$$
So $send(i, j, t, m)$ holds iff $i$ sends to $j$ message $m$ at time $t$. Likewise, let
$$recv \subseteq \Id \times \Id \times \mathbb{R} \times \Msg$$
be a relation on messages received on a channel at a time, so that $recv(i, j, t, m)$ holds iff the message $m$ received by $j$ from $i$ at time $t$.

@@

In specifying a system, a \emph{clock} module is synchronously composed (under SAL semantics for synchronous composition~\cite{SAL}) with the system. The clock module updates time to the next calendar event, and stutters otherwise.

@@

A second difficulty with model-checking fault-tolerant systems in general is that modeling faults requires adding state and nondeterminism. The minimum number of additional states that must be introduced may depend non-obviously on other aspects of the fault model, specific protocol, and system size. Such constraints lead to ``meta-model'' reasoning, such as the following, in which Rushby describes the number of data values that a particular protocol model must include to model the full range of Byzantine faults (defined later in this section):

@@

The solution to the first problem is more involved. Our solution is to introduce what we call a \emph{synchronous kibitzer} that symbolically injects faults into the model. The kibitzer decomposes the state and transitions associated with the fault model from the system itself. For the sake of concreteness in describing the synchronous kibitzer, we introduce a particular fault model, the hybrid fault model of Thambidurai and Park~\cite{hybrid}. This fault model distinguishes Byzantine, symmetric, and manifest faults. It  applies to broadcast systems in which a process is expected to broadcast the same value to multiple receivers. A \emph{Byzantine} (or \emph{arbitrary} fault is one in which a process that is intended to broadcast the same value to other processes may instead broadcast arbitrary values to different receivers (including no value or the correct value). A \emph{symmetric} fault is one in which a process may broadcast the same, but incorrect, value to other processes. Finally, a \emph{manifest} (or \emph{benign} fault is one in which a process's broadcast fault is detectable by the receivers; e.g., by performing a cyclic redundancy check (CRC) or because the value arrives outside of a predetermined window.

@@

Define a set of fault types $$\textnormal{Faults} = \{none, byz, sym, man\}.$$ As in the previous section, let $\Id \subset \mathbb{N}$ be a finite set of process indices, and let the variable $$faults: \Id \rightarrow \textnormal{Faults}$$ range over possible mappings from processes to faults.

@@

\paragraph{Transient faults.} If the $faults$ mapping is a constant, then faults are permanently but nondeterministically assigned to nodes. However, we can easily model \emph{transient faults} in which nodes are faulty temporarily by making $faults$ a state variable that is updated nondeterministically. Whether we model permanent or transient faults, a \emph{maximum fault assumption} (MFA) describes the maximum number of faults permitted in the system. The $faults$ mapping can be nondeterministically updated during execution while satisfying the MFA using a constraint such as $faults \in \{f \, | \, mfa(f) \}$, where the MFA is defined by $mfa()$.

@@

The algorithm is recursively defined:

\begin{itemize}
\item {\bf $\OMH(0)$}: $g$ broadcasts a value to each lieutenant, and the lieutenants return the value received (or $\ERR$).
\item {\bf $\OMH(m)$, $m > 0$}:
  \begin{enumerate}
  \item $g$ broadcasts a value to each lieutenant, $l$.
  \item\label{om:two} Let $l_v$ be the value received by $l \in L$ from $g$. Then for each $l$, execute $\OMH(m-1)$, assigning $l$ to be the general and $L \setminus \{l\}$ to be the lieutenants. $l$ sends $l_v$ or $R$, if $l_v = \ERR$.
  \item\label{om:three} For each lieutenant $l \in L$, remove all $\ERR$ values received in Step~\ref{om:two} from executing $\OMH(m-1)$. Compute the majority value over the remaining values, or $V$ if there is no majority. If the majority value is $R$, return $E$.
  \end{enumerate}
\end{itemize}

\noindent
In particular, $\OMH(1)$ includes two rounds of broadcast communication, one in which the general broadcasts, and one in which the lieutenants exchange their values.

$\OMH$ is designed to ensure \emph{validity} and \emph{agreement} properties under suitable hypotheses on the number and type of faults in the system. Validity states that if the general is nonfaulty, then every lieutenant outputs the value sent by the general. Agreement states that each lieutenant outputs the same value. More formally, Let $l_i, l_j$ denote the outputs of lieutenants $i, j \in L$, respectively, and let $v$ be the value the general broadcasts:
\noindent
\begin{center}
\begin{tabular}{l|l}
\begin{minipage}[t]{0.5\linewidth}
\begin{equation*}
%  \tag{Validity}
    \forall \,i. \quad l_i = v \textnormal{ (Validity)}
\end{equation*}
\end{minipage}&
\begin{minipage}[t]{0.5\linewidth}
\begin{equation*}
%  \tag{Agreement}
    \forall \,i, j. \quad l_i = l_j \textnormal{ (Agreement)}
\end{equation*}
\end{minipage}
\end{tabular}
\end{center}
%

@@

We described a hybrid fault model in Section~\ref{sec:kibitzer}. Under that fault model, agreement and validity hold if $2a+2s+b+1 \geq n$, where $n$ is the total number of nodes, $a$ is the number of Byzantine (or asymmetric) faults, $s$ is the number of symmetric faults, and $b$ is the number of benign faults. Additionally, the number of rounds $m$ must be greater or equal to the number of Byzantine faults, $a$~\cite{Lincoln-Rushby,hybrid}.

@@

Our SAL model defines seven transition systems in total: \texttt{clock}, {\texttt{source}, \texttt{relay} (parametrized over an ID), \texttt{receiver} (parametrized over an ID), \texttt{observer}, \texttt{abstractor}, and \texttt{abstract\_monitor}. The first four of these are composed asynchronously, in an intermediate system we label \texttt{system}, and share access to a global calendar consisting of event slots $(\text{message}, \text{time})$, one for each channel in the system. The \texttt{clock} transition system is responsible for updating a global variable \texttt{t} representing time according to the rules in Section~\ref{sec:calendar}.

>>>>

\newcommand{\OM}[1]{\ensuremath{\mathrm{OM}(#1)}\xspace}
\newcommand{\OMH}{\ensuremath{\mathrm{OMH}}\xspace}
\newcommand{\Id}{\ensuremath{\mathit{Id}}\xspace}
\newcommand{\Msg}{\ensuremath{\mathit{Msg}}\xspace}
\newcommand{\ERR}{\ensuremath{\mathit{ERR}}\xspace}
\newcommand{\TO}{\ensuremath{\bm{\mathrm{T}}}\xspace}
\newcommand{\tld}[1]{#1_{*}\xspace}
\newcommand{\mfa}{\ensuremath{\mathit{mfa}}\xspace}

@@

\begin{abstract}
With proof techniques like IC3 and $k$-induction, model-checking scales further
than ever before.  Still, fault-tolerant distributed systems are particularly
challenging to model-check given their large state spaces and
non-determinism. The typical approach to controlling complexity is to construct
ad-hoc abstractions of faults, message-passing, and behaviors.  However, these
abstractions come at the price of divorcing the model from its implementation
and making refactoring difficult. In this work, we present a framework for
fault-tolerant distributed system verification that combining ideas from the literature including calendar automata,
symbolic fault injection, and abstract transition systems, and then we apply the framework to model-checking various implementations of the Hybrid Oral Messages algorithm that differ in the fault model, timing model, and local node behavior. We show that despite being implementation-level models, the verifications are scalable and modular, insofar as isolated changes to an implementation require isolated changes to the model and proofs.
This work is carried out in SAL model-checker.
\end{abstract}

@@

In this paper, we present a framework for modeling fault-tolerant distributed systems, and use that framework to verify several variant implementations of the Byzantine fault-tolerant Hybrid Oral Messages algorithm (\OMH)~\cite{Lincoln-Rushby}. The framework combines ideas from the literature and folklore to build scalable and modular formal models suitable for infinite-state model-checking, and it reduces the need for ad-hoc abstractions and optimizations. In Section~\ref{sec:model}, we present the important aspects of the framework, including calendar automata, originally developed by Dutertre and Sorea~\cite{cal}, symbolic fault-injection, and abstract transition systems for verification.

@@

We use the framework to verify implementation-level models of $\OMH$ in which message passing is explicit, nodes are not forced to execute strictly synchronously, and voting is explicit. In short, the models corresponds closely with an implementation of the algorithm. In Section~\ref{sec:byz}, we first describe \OMH, then an implementation of it that uses the Boyer-Moore Fast Majority Vote algorithm (Fast~MJRT)~\cite{mjrty}. We then describe a set of modular invariants, such that the invariants only concern specific aspects of the model (e.g., faults, local node behavior, or the passage of time). The verification is interesting in its own right, as it is the first fully parametric (on the number of nodes) model-checked \emph{implementation} of the algorithm.

@@

% ------------------------------------------------------------
\section{Formal Framework}\label{sec:model}
Here we describe our formal framework specialized for fault-tolerant distributed systems. The framework draws on three principal abstractions: calendar automata, symbolic fault injection, and abstract transition systems; we describe each below.

@@

Real-time system verification in general-purpose model-checkers requires an explicit formalism of real-time progression. Trying to encode real-time clocks directly is difficult; in particular, one must avoid Zeno's paradox in which no progress is made because state transitions simply update real-valued variables by an infinite sequence of decreasing amounts whose sum is finite. To avoid this problem, Dutetre and Sorea developed \emph{calendar automata}~\cite{cal}, which is itself inspired by event calendars used in discrete-event simulation. Rather than encoding ``how much time has passed since the last event'', it encodes ``how far into the future is the next scheduled event'', and a real-valued variable representing the current time is updated to the next event time.

@@

Define a set of \emph{events} $e_0, e_1, \ldots, e_n \in E$. For now, we do not define events; intuitively, an event is a set of state variables (shortly, we will associate events with messages sent in a distributed system). When an event is \emph{enabled}, the transitions over events are enabled; otherwise, the variables stutter (maintain the same value).

@@

Let $cal$ be an event calendar and $c_i, c_j \in cal$ be calendar events. Define an ordering on calendar events such that $c_i \leq c_j$ iff $t_i \leq t_j$, and $\min(cal) = \{ c_i | \forall c_j \in cal, \, c_i \leq c_j  \}$ are the minimum elements of $cal$.

Let a transition system $\mathcal{M} = (S, I, \rightarrow)$, be a set of states $S$, a set of initial states $I \subseteq S$, and a transition relation $\rightarrow \subseteq S \times S$. We implicitly assume a set of state variables such that each $\sigma \in S$ is a total function that maps state variables to values. We distinguish two special state variables in a transition system: (1) $now \in \mathbb{R}$ denotes the current time in the state, and (2) $cal$ is an event calendar.

@@

\item \label{cal:c} In all states, if the current time is strictly less than every calendar event, then the only enabled transition is a \emph{time progress} update: $\forall \sigma \in S$, $\forall (e_i, t_i) \in \sigma(cal)$, if $\sigma(now) < t_i$, then $\forall \sigma'$ such that $\sigma \rightarrow \sigma'$, $\sigma' = \sigma[now := \min(cal)]$.

@@

Assume processes are indexed from a finite set $\Id$. A \emph{channel} from process $i$ to $j$ is an ordered pair $(i,j)$. Define a set of messages $\Msg$. Given a channel and a timeout, let $send$ be a relation on messages sent on a channel at a given time:
$$send \subseteq \Id \times \Id \times \mathbb{R} \times \Msg$$
So $send(i, j, t, m)$ holds iff $i$ sends to $j$ message $m$ at time $t$. Likewise, let
$$recv \subseteq \Id \times \Id \times \mathbb{R} \times \Msg$$
be a relation on messages received on a channel at a time, so that $recv(i, j, t, m)$ holds iff the message $m$ received by $j$ from $i$ at time $t$.

@@

@@

A second difficulty with model-checking fault-tolerant systems in general is that modeling faults requires adding state and non-determinism. The minimum number of additional states that must be introduced may depend non-obviously on other aspects of the fault model, specific protocol, and system size. Such constraints lead to ``meta-model'' reasoning, such as the following, in which Rushby describes the number of data values that a particular protocol model must include to model the full range of Byzantine faults (defined later in this section):

@@

The solution to the first problem is more involved. Our solution is to introduce what we call a \emph{synchronous kibitzer} that symbolically injects faults into the model. The kibitzer decomposes the state and transitions associated with the fault model from the system itself. For the sake of concreteness in describing the synchronous kibitzer, we introduce a particular fault model, the hybrid fault model of Thambidurai and Park~\cite{hybrid}. This fault model distinguishes Byzantine, symmetric, and manifest faults. It  applies to broadcast systems in which a process is expected to broadcast the same value to multiple receivers. A \emph{Byzantine} (or \emph{arbitrary}) fault is one in which a process that is intended to broadcast the same value to other processes may instead broadcast arbitrary values to different receivers (including no value or the correct value). A \emph{symmetric} fault is one in which a process may broadcast the same, but incorrect, value to other processes. Finally, a \emph{manifest} (or \emph{benign}) fault is one in which a process's broadcast fault is detectable by the receivers; e.g., by performing a cyclic redundancy check (CRC) or because the value arrives outside of a predetermined window.

@@

Define a set of fault types $$\textnormal{Faults} = \{none, byz, sym, man\}.$$ As in the previous section, let $\Id$ be a finite set of process indices, and let the variable $$faults: \Id \rightarrow \textnormal{Faults}$$ range over possible mappings from processes to faults.

@@

If the $faults$ mapping is a constant, then faults are permanently but non-deterministically assigned to nodes. However, we can easily model \emph{transient faults} in which nodes are faulty temporarily by making $faults$ a state variable that is updated non-deterministically. Whether we model permanent or transient faults, a \emph{maximum fault assumption} (MFA) describes the maximum number of faults permitted in the system. The $faults$ mapping can be non-deterministically updated during execution while satisfying the MFA using a constraint such as $faults \in \{f \, | \, \mfa(f) \}$, where the MFA is defined by the function $\mfa$.

@@

The algorithm is recursively defined for $m \ge 0$:

\begin{itemize}
\item {\bf $\OMH(0)$}: $g$ broadcasts a value to each lieutenant and the lieutenants return the value received (or $\ERR$).
\item {\bf $\OMH(m)$, $m > 0$}:
  \begin{enumerate}
  \item $g$ broadcasts a value to each lieutenant, $l$.
  \item\label{om:two} Let $l_v$ be the value received by $l \in L$ from $g$. Then for each $l$, execute $\OMH(m-1)$, assigning $l$ to be the general and $L \setminus \{l\}$ to be the lieutenants. $l$ sends $l_v$, or $R$ if $l_v = \ERR$.
  \item\label{om:three} For each lieutenant $l \in L$, remove all $\ERR$ values received in Step~\ref{om:two} from executing $\OMH(m-1)$. Compute the majority value over the remaining values, or $V$ if there is no majority. If the majority value is $R$, return $E$.
  \end{enumerate}
\end{itemize}

\noindent
In particular, $\OMH(1)$ includes two rounds of broadcast communication: one in which the general broadcasts, and one in which the lieutenants exchange their values.

$\OMH$ is designed to ensure \emph{validity} and \emph{agreement} properties under suitable hypotheses on the number and type of faults in the system. Validity states that if the general is nonfaulty, then every lieutenant outputs the value sent by the general. Agreement states that each lieutenant outputs the same value. More formally, Let $l_i, l_j$ denote the outputs of lieutenants $i, j \in L$, respectively, and let $v$ be the value the general broadcasts:
\vspace{-0.4cm}
\noindent
\begin{center}
\begin{tabular}{l|l}
\begin{minipage}[t]{0.5\linewidth}
\begin{equation*}
%  \tag{Validity}
    \forall \,i. \quad l_i = v \textnormal{ (Validity)}
\end{equation*}
\end{minipage}&
\begin{minipage}[t]{0.5\linewidth}
\begin{equation*}
%  \tag{Agreement}
    \forall \,i, j. \quad l_i = l_j \textnormal{ (Agreement)}
\end{equation*}
\end{minipage}
\end{tabular}
\end{center}
%

@@

We described a hybrid fault model in Section~\ref{sec:kibitzer}. Under that fault model, validity and agreement hold if $2a+2s+b+1 \geq n$, where $n$ is the total number of nodes, $a$ is the number of Byzantine (or asymmetric) faults, $s$ is the number of symmetric faults, and $b$ is the number of benign faults. Additionally, the number of rounds $m$ must be greater or equal to the number of Byzantine faults, $a$~\cite{Lincoln-Rushby,hybrid}.

@@

Our SAL model defines seven transition systems in total: \texttt{clock}, {\texttt{source}, \texttt{relay} (parametrized over an ID), \texttt{receiver} (parametrized over an ID), \texttt{observer}, \texttt{abstractor}, and \texttt{abstract\_monitor}. The first four of these are composed asynchronously, in an intermediate system we label \texttt{system}, and share access to a global calendar consisting of event slots $(\text{message}, \text{time})$, one for each channel in the system. The \texttt{clock} transition system is responsible for updating a global variable \texttt{t} (called $now$ in Section~\ref{sec:calendar}) representing time according to the rules for calendar automata.
